---
title: "(s)PLS and (s)PLS-DA"
author: "Warasinee Mujchariyakul"
output: md_document
date: "2025-09-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load R packages
```{r}
library(mixOmics)
library(tidyverse)
set.seed(123) # for reproducibility
```


## Import all omics
```{r}
Strep_trans_data <- read.csv("/Users/wmujchariyak/Desktop/Strep_transcriptome_data.csv", header = TRUE) %>% column_to_rownames(var = "X")
Strep_prot_data <- read.csv("/Users/wmujchariyak/Desktop/Strep_proteome_data.csv", header = TRUE) %>% column_to_rownames(var = "X")
Strep_met_GC_data <- read.csv("/Users/wmujchariyak/Desktop/Strep_metabolome_GC_data.csv", header = TRUE) %>% column_to_rownames(var = "X")
```


## Data for MixOmics 
```{r}
data <- list(Transcript = Strep_trans_data, 
             Protein = Strep_prot_data, 
             Metabolite_GC = Strep_met_GC_data)
lapply(data, dim) # check their dimensions (24 rows)


Y <- factor(rep(c("RPMI", "SERUM"), 30)) # set the response variable as the Y df (based on conditions)
Y2 <- factor(rep(c("5448_1", "HKU419","PS003","PS006","SP444"), each = 12)) # set the response variable as the Y df  (based on strains)
```


# Pairwise PLS Comparisons (ncomp=1) 
** Calculate correlation between omics datasets which can be beneficial for setting of design matrix 
```{r}
# generate pairwise PLS models 
pls1 <- pls(data$Transcript, data$Protein, ncomp = 1) 
pls2 <- pls(data$Protein, data$Metabolite_GC, ncomp = 1)
pls3 <- pls(data$Transcript, data$Metabolite_GC, ncomp = 1)

# calculate correlation 
cor(pls1$variates$X, pls1$variates$Y) #cor trans vs prot = 0.9319566 
cor(pls2$variates$X, pls2$variates$Y) #cor prot vs met_GC = 0.9127586
cor(pls3$variates$X, pls3$variates$Y) #cor trans vs met_GC = 0.9343444
```


## sPLS-DA: based on conditions
** An sPLS-DA analysis will help refine the sample clusters and select a small subset of variables relevant to discriminate each class. 
```{r}
## Transcript 
# Load the data
X.trans <- data$Transcript
dim(X.trans); length(Y)

# Initial model
plsda.trans <- plsda(X.trans,Y, ncomp = 10)

# Tuning number of components
perf.plsda.trans <- perf(plsda.trans, validation = 'loo', 
                  progressBar = FALSE,  # Set to TRUE to track progress
                  nrepeat = 1)    

head(perf.plsda.trans$error.rate) 
plot(perf.plsda.trans, sd = TRUE, legend.position = 'horizontal')
# The error rate decreases and reaches a minimum for ncomp = 2 for the max.dist distance. These parameters will be included in further analyses.

# Final model
final.plsda.trans <- plsda(X.trans,Y, ncomp = 2)

plotIndiv_trans <- plotIndiv(final.plsda.trans, ind.names = T, legend=TRUE,
          comp=c(1,2), ellipse = TRUE, 
          title = 'PLS-DA on Transcriptome comp 1-2',
          X.label = 'PLS-DA comp 1', Y.label = 'PLS-DA comp 2')

######################################
## Proteomics 
# Load the data
X.prot <- data$Protein
dim(X.prot); length(Y)

# Initial model
plsda.prot <- plsda(X.prot,Y, ncomp = 10)

# Tuning number of components
perf.plsda.prot <- perf(plsda.prot, validation = 'loo', 
                  progressBar = FALSE,  # Set to TRUE to track progress
                  nrepeat = 1)    

head(perf.plsda.prot$error.rate) 
plot(perf.plsda.prot, sd = TRUE, legend.position = 'horizontal')
# The error rate decreases and reaches a minimum for ncomp = 1 for the max.dist distance. These parameters will be included in further analyses.

# Final model
final.plsda.prot <- plsda(X.prot,Y, ncomp = 2) # we use ncomp =2 so that we can visualize with plotIndiv

plotIndiv_prot <- plotIndiv(final.plsda.prot, ind.names = T, legend=TRUE,
          comp=c(1,2), ellipse = TRUE, 
          title = 'PLS-DA on Proteome comp 1-2',
          X.label = 'PLS-DA comp 1', Y.label = 'PLS-DA comp 2')

######################################
## Met - sPLS-DA
# Load the data
X.met <- data$Metabolite_GC
dim(X.met); length(Y)

# Initial model
plsda.met <- plsda(X.met,Y, ncomp = 10)

# Tuning number of components
perf.plsda.met <- perf(plsda.met, validation = 'loo', 
                  progressBar = FALSE,  # Set to TRUE to track progress
                  nrepeat = 1)    

head(perf.plsda.met$error.rate) 
plot(perf.plsda.met, sd = TRUE, legend.position = 'horizontal')
# The error rate decreases and reaches a minimum for ncomp = 1 for the max.dist distance. These parameters will be included in further analyses.

# Final model
final.plsda.met <- plsda(X.met,Y, ncomp = 2) # we use ncomp =2 so that we can visualize with plotIndiv

plotIndiv_met <- plotIndiv(final.plsda.met, ind.names = T, legend=TRUE,
          comp=c(1,2), ellipse = TRUE, 
          title = 'PLS-DA on Metabolome comp 1-2',
          X.label = 'PLS-DA comp 1', Y.label = 'PLS-DA comp 2')


```


## sPLS-DA: based on strains
```{r}
## Transcript 
# Load the data
X.trans <- data$Transcript
dim(X.trans); length(Y2)

# Initial model
plsda.trans.2 <- plsda(X.trans,Y2, ncomp = 10)

# Tuning number of components
perf.plsda.trans.2 <- perf(plsda.trans.2, validation = 'loo', 
                  progressBar = FALSE,  # Set to TRUE to track progress
                  nrepeat = 1)    

head(perf.plsda.trans.2$error.rate) 
plot(perf.plsda.trans.2, sd = TRUE, legend.position = 'horizontal')
# The error rate decreases and reaches a minimum for ncomp = 4 for the max.dist distance. These parameters will be included in further analyses.

# Final model
final.plsda.trans.2 <- plsda(X.trans,Y2, ncomp = 4)

plotIndiv_trans.2 <- plotIndiv(final.plsda.trans.2, ind.names = F, legend=TRUE,
          comp=c(1,2), ellipse = TRUE, 
          title = 'PLS-DA on Transcriptome comp 1-2',
          X.label = 'PLS-DA comp 1', Y.label = 'PLS-DA comp 2')

######################################
## Proteomics 
# Load the data
X.prot <- data$Protein
dim(X.prot); length(Y2)

# Initial model
plsda.prot.2 <- plsda(X.prot,Y2, ncomp = 10)

# Tuning number of components
perf.plsda.prot.2 <- perf(plsda.prot.2, validation = 'loo', 
                  progressBar = FALSE,  # Set to TRUE to track progress
                  nrepeat = 1)    

head(perf.plsda.prot.2$error.rate) 
plot(perf.plsda.prot.2, sd = TRUE, legend.position = 'horizontal')
# The error rate decreases and reaches a minimum for ncomp = 5 for the max.dist distance. These parameters will be included in further analyses.

# Final model
final.plsda.prot.2 <- plsda(X.prot,Y2, ncomp = 5) 

plotIndiv_prot.2 <- plotIndiv(final.plsda.prot.2, ind.names = F, legend=TRUE,
          comp=c(1,2), ellipse = TRUE, 
          title = 'PLS-DA on Proteome comp 1-2',
          X.label = 'PLS-DA comp 1', Y.label = 'PLS-DA comp 2')

######################################
## Met - sPLS-DA
# Load the data
X.met <- data$Metabolite_GC
dim(X.met); length(Y2)

# Initial model
plsda.met.2 <- plsda(X.met,Y2, ncomp = 10)

# Tuning number of components
perf.plsda.met.2 <- perf(plsda.met.2, validation = 'loo', 
                  progressBar = FALSE,  # Set to TRUE to track progress
                  nrepeat = 1)    

head(perf.plsda.met.2$error.rate) 
plot(perf.plsda.met.2, sd = TRUE, legend.position = 'horizontal')
# The error rate decreases and reaches a minimum for ncomp = 5 for the max.dist distance. These parameters will be included in further analyses.

# Final model
final.plsda.met.2 <- plsda(X.met,Y2, ncomp = 5) # we use ncomp =2 so that we can visualize with plotIndiv

plotIndiv_met.2 <- plotIndiv(final.plsda.met.2, ind.names = F, legend=TRUE,
          comp=c(1,2), ellipse = TRUE, 
          title = 'PLS-DA on Metabolome comp 1-2',
          X.label = 'PLS-DA comp 1', Y.label = 'PLS-DA comp 2')

```
## Save 
```{r}
#pdf("/Users/wmujchariyak/Desktop/Strep_plotIndiv_plsda_trans.pdf", height = 5, width = 8)
#plotIndiv_trans
#dev.off()

#pdf("/Users/wmujchariyak/Desktop/Strep_plotIndiv_plsda_prot.pdf", height = 5, width = 8)
#plotIndiv_prot
#dev.off()

#pdf("/Users/wmujchariyak/Desktop/Strep_plotIndiv_plsda_met_GC.pdf", height = 5, width = 8)
#plotIndiv_met
#dev.off()

#pdf("/Users/wmujchariyak/Desktop/Strep_plotIndiv_plsda_trans2.pdf", height = 5, width = 8)
#plotIndiv_trans.2
#dev.off()

#pdf("/Users/wmujchariyak/Desktop/Strep_plotIndiv_plsda_prot2.pdf", height = 5, width = 8)
#plotIndiv_prot.2
#dev.off()

#pdf("/Users/wmujchariyak/Desktop/Strep_plotIndiv_plsda_met_GC2.pdf", height = 5, width = 8)
#plotIndiv_met.2
#dev.off()
```

