---
title: "DIABLO - Identification of Signature Molecules of *Streptococcus pyogenes* Serum Responses"
author: "Warasinee Mujchariyakul"
output: md_document
date: "2025-09-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load R packages
```{r}
library(mixOmics)
library(tidyverse)
library(igraph)
set.seed(123) # for reproducibility
```

## Import all omics
```{r}
Strep_trans_data <- read.csv("/Users/wmujchariyak/Desktop/Strep_transcriptome_data.csv", header = TRUE) %>% column_to_rownames(var = "X")
Strep_prot_data <- read.csv("/Users/wmujchariyak/Desktop/Strep_proteome_data.csv", header = TRUE) %>% column_to_rownames(var = "X")
Strep_met_GC_data <- read.csv("/Users/wmujchariyak/Desktop/Strep_metabolome_GC_data.csv", header = TRUE) %>% column_to_rownames(var = "X")
```

## Data for MixOmics (DIABLO)
```{r}
data <- list(Transcript = Strep_trans_data, 
             Protein = Strep_prot_data, 
             Metabolite_GC = Strep_met_GC_data)
lapply(data, dim) # check their dimensions (60 rows)


Y <- factor(rep(c("RPMI", "SERUM"), 30)) # set the response variable as the Y df
summary(Y)
```

## Initial DIABLO model 
** With the pairwise PLS comparison results (cor > 0.9), all omics datasets are highly connected.
Therefore, we can adjust "design matrix" to 0.5 to balance power between identifying correlation features across omics and achieving classification accuracy.  
** 1 = maximize relationship enforced (between blocks), 0 = no relationship enforced (discrimination focus)
** diagonal is set to zero so there is no self linked (in the same block). 
```{r}
design = matrix(0.5, ncol = length(data), nrow = length(data), 
                dimnames = list(names(data), names(data)))
diag(design) = 0 # set diagonal to 0s
design
```

## Tuning parameters
** [LOO-CV issue](https://mixomics-users.discourse.group/t/loo-cv-outputs-na-in-choice-ncomp/145/3) - you can look at the perf plot and decide on the number of components (Here, ncomp = 1).  
```{r}
# Form basic DIABLO model with  an arbitrarily high number of components (ncomp = 5)
basic.diablo.model = block.splsda(X = data, Y = Y, ncomp = 5, design = design)

# Tuning the number of components ("loo" = leave-one-out cross-validation, for small sample sizes)
perf.diablo = perf(basic.diablo.model, validation = 'loo', nrepeat = 1) 
plot(perf.diablo) # plot output of tuning
# Set the optimal ncomp value
ncomp = perf.diablo$choice.ncomp$WeightedVote["Overall.BER", "centroids.dist"] 
# Show the optimal choice for ncomp for each dist metric
perf.diablo$choice.ncomp$WeightedVote 
```

```{r}
# Check output - basic DIABLO model
plotIndiv(basic.diablo.model, ind.names=FALSE, legend=TRUE)
plotLoadings(basic.diablo.model)
```


## Tuning the number of features 
** This may take some time to run (~ 5 min)
```{r}
# Set grid of values for each component to test
test.keepX = list (Transcript = seq(20,100,20), 
                   Protein = seq(20,100,20),
                   Metabolite_GC = seq(10,100,10)) 

# Note: seq(10,100,10) = start from 10 and increase by 10 until 100

# Run the feature selection tuning (ncomp = 1 - from tuning the number of components)
tune.SA = tune.block.splsda(X = data, Y = Y, ncomp = 1, 
                              test.keepX = test.keepX, design = design,
                              validation = 'Mfold', folds = 5, nrepeat = 10,
                              dist = "centroids.dist")
# Set the optimal values of features to retain
list.keepX = tune.SA$choice.keepX 
list.keepX
```


## Final DIABLO model
```{r}
# Final model (use arbitrary number for # variables from our preliminary results in "Tuning the number of features")

list.keepX <-  list (Transcript =  20, 
                    Protein =  20,
                    Metabolite_GC = 10) 

# Set the optimised DIABLO model
final.diablo.model = block.splsda(X = data, Y = Y, ncomp = 1, 
                                  keepX = list.keepX, design = design)

final.diablo.model$design # design matrix for the final model
```

## AUC for final model
```{r}
auc.diablo.trans.com <- auroc(final.diablo.model, roc.block = "Transcript", roc.comp = 1,
                        print = FALSE)
auc.diablo.prot.com <- auroc(final.diablo.model, roc.block = "Protein", roc.comp = 1,
                        print = FALSE)
auc.diablo.met.GC.com <- auroc(final.diablo.model, roc.block = "Metabolite_GC", roc.comp = ,
                        print = FALSE)
```

## Extract data from final DIABLO model 
```{r}
# Variables selected on component 1
trans_var <- selectVar(final.diablo.model, block = 'Transcript', comp = 1)
prot_var <- selectVar(final.diablo.model, block = 'Protein', comp = 1)
met_GC_var <- selectVar(final.diablo.model, block = 'Metabolite_GC', comp = 1)

# Correlation matrix from the circos plot
corMat <- circosPlot(final.diablo.model, cutoff = 0.7)
corMat2 <- as.data.frame(corMat)
```

## Data visualization
```{r}
# Sample plots
plotDiablo(final.diablo.model, ncomp = 1)
```

```{r fig.height=6, fig.width=8}
# Plot Loadings
plotLoadings(final.diablo.model, comp = 1, contrib = 'max', method = 'median', size.name = 0.65)
```


```{r fig.height=9.5, fig.width=9.5}
# Circos plot 
circosPlot(final.diablo.model, cutoff = 0.9, comp = 1, line = TRUE, 
           color.blocks = c('darkorchid', 'brown1', 'lightgreen'),
           color.cor = c("chocolate3","grey20"), size.labels = 1.2, size.variables = 0.68, size.legend = 1)
```

## Save 
```{r fig.height=10, fig.width=10}
# Save data for further visualization in Cytoscape
myNetwork <- network(final.diablo.model, blocks = c(1,2,3), cutoff = 0.9) 
#write_graph(myNetwork$gR, file = "/Users/wmujchariyak/Desktop/myNetwork_conserved_Strep.gml", format = "gml")
```

## loading values + Gene name 
```{r}
# We used loading values for ranking important viables
loadings_data <- final.diablo.model$loadings 

loadings_data_trans <- loadings_data$Transcript
loadings_data_prot <- loadings_data$Protein
loadings_data_met <- loadings_data$Metabolite_GC

# Import all entities fold change results (all omics combined)
all_DE.df <- readRDS("/Users/wmujchariyak/Desktop/All_entity_fold_change_long_format.rds")
Strep_DE.df <- readRDS("/Users/wmujchariyak/Desktop/All_entity_fold_change_long_format.rds") %>% filter(species == "Streptococcus pyogenes") 
Strep_DE.df.trans <- Strep_DE.df %>% filter(experiment == "Transcriptomics RG") %>% dplyr::select(core_entity_id, core_consensus_annotation,is_species_core) %>% unique()
Strep_DE.df.prot <- Strep_DE.df %>% filter(experiment == "Proteomics MS1 DDA") %>% dplyr::select(core_entity_id,core_consensus_annotation,is_species_core) %>% unique()
Strep_DE.df.met <- Strep_DE.df %>% filter(experiment == "Metabolomics GC-MS") %>% dplyr::select(core_entity_id, additional_id,is_species_core)%>% unique()

loadings_data_trans_2 <- loadings_data_trans %>% as.data.frame(.) %>% filter(comp1 != 0) %>% rownames_to_column(., var = "core_entity_id") %>% left_join(., Strep_DE.df.trans, by = "core_entity_id" )
loadings_data_prot_2 <- loadings_data_prot %>% as.data.frame(.) %>% filter(comp1 != 0) %>% rownames_to_column(., var = "core_entity_id") %>% left_join(., Strep_DE.df.prot, by = "core_entity_id" )
loadings_data_met_2 <- loadings_data_met %>% as.data.frame(.) %>% filter(comp1 != 0) %>% rownames_to_column(., var = "core_entity_id") %>% left_join(., Strep_DE.df.met, by = "core_entity_id" )

#save
#write.table(loadings_data_trans_2, file = "/Users/wmujchariyak/Desktop/strep_loadings_data_trans_2.tsv", sep ="\t", row.names = F)
#write.table(loadings_data_prot_2, file = "/Users/wmujchariyak/Desktop/strep_loadings_data_prot_2.tsv", sep ="\t", row.names = F)
#write.table(loadings_data_met_2, file = "/Users/wmujchariyak/Desktop/strep_loadings_data_met_2.tsv", sep ="\t", row.names = F)
```

