---
title: "DIABLO results for network analysis"
author: "Warasinee Mujchariyakul"
output: md_document
date: "2025-06-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load R packages
```{r}
library(mixOmics)
library(tidyverse)
library(igraph)
library(DOSE)
library(clusterProfiler)
library(RCy3)
library(ggraph)

set.seed(123) # for reproducibility
```

## Import all omics
```{r}
Strep_trans_data <- read.csv("/Users/wmujchariyak/Desktop/Strep_transcriptome_data.csv", header = TRUE) %>% column_to_rownames(var = "X")
Strep_prot_data <- read.csv("/Users/wmujchariyak/Desktop/Strep_proteome_data.csv", header = TRUE) %>% column_to_rownames(var = "X")
Strep_met_GC_data <- read.csv("/Users/wmujchariyak/Desktop/Strep_metabolome_GC_data.csv", header = TRUE) %>% column_to_rownames(var = "X")
```

## Data for MixOmics (DIABLO)
```{r}
data <- list(Transcript = Strep_trans_data, 
             Protein = Strep_prot_data, 
             Metabolite_GC = Strep_met_GC_data)
lapply(data, dim) # check their dimensions (60 rows)


Y <- factor(rep(c("RPMI", "SERUM"), 30)) # set the response variable as the Y df
summary(Y)
```

## Initial DIABLO model 
```{r}
design = matrix(0.1, ncol = length(data), nrow = length(data), 
                dimnames = list(names(data), names(data)))
diag(design) = 0 # set diagonal to 0s
design
```

## Tuning parameters
** [LOO-CV issue](https://mixomics-users.discourse.group/t/loo-cv-outputs-na-in-choice-ncomp/145/3) - you can look at the perf plot and decide on the number of components (Here, ncomp = 1).  
```{r}
# Form basic DIABLO model with  an arbitrarily high number of components (ncomp = 5)
plsda.diablo <- block.plsda(data, Y, ncomp = 5, design = design)

set.seed(123) # for reproducibility
# Tuning the number of components ("loo" = leave-one-out cross-validation, for small sample sizes)
perf.plsda.diablo = perf(plsda.diablo, validation = 'loo', nrepeat = 1) 

#perf.plsda.diablo$error.rate  # Lists the different types of error rates
# Plot of the error rates based on weighted vote
plot(perf.plsda.diablo) # plot output of tuning
```

## Final model - block.plsda (This method is different from block.splsda)
```{r}
final.plsda.diablo <- block.plsda(data, Y, ncomp = 1, design = design)
```

## Extract loading values (for ranking importanct viables)
```{r}
loadings_data <- final.plsda.diablo$loadings 

loadings_data_trans <- loadings_data$Transcript
loadings_data_prot <- loadings_data$Protein
loadings_data_met <- loadings_data$Metabolite_GC
```

## Import all omics (annotation)
```{r}
# Import all entities fold change results (all omics combined)
all_DE.df <- readRDS("/Users/wmujchariyak/Desktop/All_entity_fold_change_long_format.rds")
all_DE.df.short <- all_DE.df %>%  dplyr::select(entity_id, core_entity_id) %>% unique()

# Import all functional annotations 
all.annot <- readRDS("/Users/wmujchariyak/Desktop/All_funct_annot_clean_noduplicate.rda")
```

## Set parameters
```{r}
# For ORA and GSEA
annot_origin <- unique(all.annot$fterm_origin) %>% as.character() # Use all annotations
```

## GSEA analysis  
```{r}
## Prepare Transcriptome data
my_species <- "Streptococcus pyogenes"
my_omics <- "Transcriptomics RG"

all_DE.res <- all_DE.df %>%
  filter(species == my_species) %>%
  filter(experiment == my_omics)

# Annotation - add core_entity_id 
if (grepl("Metabo", my_omics)){
  my.annot <- all.annot %>%
    filter(entity_type == "metabolite")
} else {
  my.annot <- all.annot %>%
    filter(entity_type != "metabolite")
}


# Loading values 
GSEA.df.trans <- loadings_data_trans %>% as.data.frame(.) %>% 
                 rownames_to_column(., var = "core_entity_id") %>% 
                 rename(loadings = comp1)

# Get background gene list  (all variables from MixOmics)
bglist.trans <- GSEA.df.trans$core_entity_id

term_to_gene.df.trans <- my.annot %>%
    filter(fterm_origin %in% annot_origin) %>% 
    left_join(., all_DE.df.short, by = "entity_id") %>% 
    filter(core_entity_id %in% bglist.trans) %>%
    distinct(fterm_id, core_entity_id) %>%
    ungroup()

term_to_name.df.trans <- my.annot %>%
    filter(fterm_origin %in% annot_origin) %>%
    left_join(., all_DE.df.short, by = "entity_id") %>% 
    filter(core_entity_id %in% bglist.trans) %>%
    dplyr::select(fterm_id, fterm_name) %>%
    distinct(fterm_id, fterm_name) %>%
    ungroup()
#note: Warning: Detected an unexpected many-to-many relationship between `x` and `y`.

###########################################################
## GSEA analysis

glist.GSEA.trans <- GSEA.df.trans$loadings
names(glist.GSEA.trans) <- GSEA.df.trans$core_entity_id
glist.GSEA.trans <- sort(glist.GSEA.trans, decreasing = T)

GSEA.output.trans <- GSEA(geneList = glist.GSEA.trans, 
               TERM2GENE = term_to_gene.df.trans,
               TERM2NAME = term_to_name.df.trans,
               minGSSize = 5,
               maxGSSize = length(bglist.trans)/2,
               pvalueCutoff = 1,
               pAdjustMethod = "fdr",
               nPermSimple = 100000,  # Increase the number of permutations
               eps = 0  # Allow more precise p-value estimation
               )
```

# Save 
```{r}
#saveRDS(GSEA.output.trans, file = "/Users/wmujchariyak/Desktop/Strep.DIABLO.GSEA.output.trans.rds")
```

## Collect node and edge - DIABLO.GSEA.KEGG.sig
```{r}
# 1. Prepare data
Trans.DIABLO.GSEA.KEGG.sig <- clusterProfiler::filter(GSEA.output.trans, p.adjust < 0.05, grepl("^map", ID))

# 2. Extract node and edge data from Trans.GSEA.KEGG 
# 2.1 Calculate pairwise similarity (of the enriched terms using Jaccard’s similarity index)
      # Check added similarity matrix in the termsim slot of enrichment result 
      # For example, View(Trans.DIABLO.GSEA.KEGG.sig_2@termsim)
Trans.DIABLO.GSEA.KEGG.sig_2 <- enrichplot::pairwise_termsim(Trans.DIABLO.GSEA.KEGG.sig, method = "JC")  

# 2.2 Create Enrichment map 
emap_Trans.DIABLO.GSEA.KEGG.sig <- enrichplot::emapplot(Trans.DIABLO.GSEA.KEGG.sig_2)

# 2.3 Function to convert enrichment map to dataframe
emap_to_network_df <- function(emap, what = "edges"){
  if (what == "vertices"){
    df <- emap$data %>% 
      attributes %>% # provide data attributes in list
      .$graph %>% # access an attribute named graph (which is in igraph format containing vertices (node) and edges) 
      igraph::as_data_frame(what = "vertices") %>% # convert igraph nodes (vertices) to dataframe
      rename(gene_set_size = "size") %>%
      rename(p.adjust = "color") %>%
      select(name, gene_set_size, p.adjust)
    return(df)
  } else if (what == "edges"){
    df <- emap$data %>%
      attributes %>%
      .$graph %>%
      igraph::as_data_frame(what = "edges") %>%
      rename(similarity = "weight") %>%
      select(from, to, similarity)
    return(df)
  } else {
    return(print("error what should be set to edges or nodes"))
  }
}

# 2.4 Prepare edges df and node df of 1 strain (Please review 5_Node_Edge_data_Strep_220425.md for multiple strains)
all_edge_DIABLO_trans_GSEA_KEGG <- emap_to_network_df(emap_Trans.DIABLO.GSEA.KEGG.sig) %>%
  dplyr::rowwise() %>% # allows you to compute on a data frame a row-at-a-time
  dplyr::mutate(edge = paste0(sort(c(from, to)), collapse = "#")) %>%
  dplyr::group_by(edge) %>%
  dplyr::summarize(nb_strain_with_edge = n(),
                   #similarity = paste0(similarity, collapse = ",")) %>%
                   similarity = mean(similarity)) %>%
  ungroup() %>%
  tidyr::separate(edge, sep = "#", into = c("to", "from")) %>%
  dplyr::select(from, to, nb_strain_with_edge, similarity)


all_node_DIABLO_trans_GSEA_KEGG <- emap_to_network_df(emap_Trans.DIABLO.GSEA.KEGG.sig, what = "vertices")  %>%
  dplyr::group_by(name) %>%
  dplyr::summarize(nb_strain_with_node = n(),
                   #gene_set_size = paste0(gene_set_size, collapse = ","),
                   gene_set_size = mean(gene_set_size),
                   p.adjust =paste0(p.adjust, collapse = ",")) %>%
                   ungroup()
```

# save 
```{r}
#write.table(all_edge_DIABLO_trans_GSEA_KEGG, "/Users/wmujchariyak/Desktop/Strep_all_edge_DIABLO_trans_GSEA_KEGG.tsv", sep = "\t", row.names = F)
#write.table(all_node_DIABLO_trans_GSEA_KEGG, "/Users/wmujchariyak/Desktop/Strep_all_node_DIABLO_trans_GSEA_KEGG.tsv", sep = "\t", row.names = F)
```

## Network ananlysis to visualize interactions of enriched pathways 
```{r}
# Example for only transcriptomics (GSEA - KEGG)

# 1. Create igraph graphs from data frames
# Check: class(ig_trans_GSEA_KEGG)
ig_trans_GSEA_KEGG <- igraph::graph_from_data_frame(d=all_edge_DIABLO_trans_GSEA_KEGG, vertices=all_node_DIABLO_trans_GSEA_KEGG, directed = FALSE)

# 2. Add labels to the nodes 
tg_trans_GSEA_KEGG <- tidygraph::as_tbl_graph(ig_trans_GSEA_KEGG) %>% 
  tidygraph::activate(nodes) %>% 
  dplyr::mutate(label=name)

# 3. Plot the network
# set seed (so that the exact same network graph is created every time)
set.seed(12345)

# 3.1 Create a graph using the 'tg' data frame with the Fruchterman-Reingold layout
plot_tg_trans_GSEA_KEGG_3 <- tg_trans_GSEA_KEGG %>%
  ggraph::ggraph(layout = "fr") +
  
  # Add arcs for edges with various aesthetics
  geom_edge_arc(colour = "gray50",
                lineend = "round",
                strength = .1,
                aes(edge_width = all_edge_DIABLO_trans_GSEA_KEGG$nb_strain_with_edge,
                    alpha = all_edge_DIABLO_trans_GSEA_KEGG$nb_strain_with_edge)) +
  
  # Add points for nodes with size and color (for 1 strain)
  ggraph::geom_node_point(size = 5,
                          aes(color = all_node_DIABLO_trans_GSEA_KEGG$nb_strain_with_node)) +
  
  # Add edge of nodes on the next layer (shape = 21)
  ggraph::geom_node_point(size = 5, shape = 21) + 
  
  # Add text labels for nodes with various aesthetics
  geom_node_text(aes(label = name), 
                 repel = TRUE, 
                 point.padding = unit(0.2, "lines"), 
                 colour = "gray10") +
  
  # Change color palette with custom colors
  scale_color_gradientn(
    colors ="#D73027") +
  
  # Adjust edge width and alpha scales
  scale_edge_width(range = c(0.5, 2.5)) +
  scale_edge_alpha(range = c(0.2, 0.5)) +
  
  # Set graph background color to white
  theme_graph(background = "white") +
  
  # Adjust legend position to the top
  theme(legend.position = "bottom", 
        # suppress legend title
        legend.title = element_blank()) 
```

```{r fig.height=5, fig.width=10}
# 3.2 View the plot 
plot_tg_trans_GSEA_KEGG_3 # with fig.height=8, fig.width=20
```



